---
title: Hatchery Inseason Forecasts
author: Thomas Buehrens (tbuehrens@dfw.wa.gov) 
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

***

Last Updated `r format(Sys.time(), '%m/%d/%Y')`.

***


# Overview
This script fits two regression models to in-season hatchery return data to predict total return

## Setup
All analyses require R software [**(link)**](https://cran.r-project.org/) (v3.4.3) for data retrieval, data processing, and summarizing model results. Here we configure R to perform our analysis and generate our outputs
```{r set_options, echo = TRUE, message = FALSE}
options(width = 100)
knitr::opts_chunk$set(message = FALSE)
set.seed(123)
```

We also need a couple of helper functions which we will define
```{r load_funcs, message = FALSE, warning = FALSE,results = "hide"}
#function to install or load packages
install_or_load_pack <- function(pack){
  create.pkg <- pack[!(pack %in% installed.packages()[, "Package"])]
  if (length(create.pkg))
    install.packages(create.pkg, dependencies = TRUE)
  sapply(pack, require, character.only = TRUE)
}

#function to make seasonal timeseries like hatchery returns
make_data<-function(S,min_yr,Y,Y_resid,n,phi,theta,proc_sd){
  mu = -3 + (0.2 - ((0.2/S) * 1:S)) * (1:S) 
  series <-as_tibble(sim.ssarima(orders = list(ar=c(1,1),i=c(0,0),ma=c(1,1)),
                        lags=c(1,S),
                        obs=n, 
                        AR=phi, 
                        MA=theta,
                        constant = 0,
                        initial = rnorm(S+1, 0 ,proc_sd*2),
                        randomizer = "rnorm",
                        n=,
                        mean=0,
                        sd=proc_sd)
                        $data)%>%
    mutate(year = sort(rep(1:Y,S))+minyr-1, yday = rep(1:S,Y), date = date_build(year=year, month = ifelse(yday>31,2,1), day = ifelse(yday>31,yday-31,yday)))%>%
    dplyr::rename(abundance= `Series 1`)
  
  series<-series%>%
    mutate(abundance =rpois(n=n,lambda=exp(series$abundance + mu[series$yday] + int + Y_resid[series$year-minyr+1])))
}

#prediction intervals GAM
gam_pred_ints<-function(fit_gam,new_dat,quants){
  beta <- coef(fit_gam)
  V <- vcov(fit_gam)
  num_beta_vecs <- 10000
  Cv <- chol(V)
  set.seed(1)
  nus <- rnorm(num_beta_vecs * length(beta))
  beta_sims <- beta + t(Cv) %*% matrix(nus, nrow = length(beta), ncol = num_beta_vecs)
  covar_sim <- predict(fit_gam, newdata = new_dat, type = "lpmatrix")
  linpred_sim <- covar_sim %*% beta_sims
  invlink <- function(x) exp(x)
  exp_val_sim <- invlink(linpred_sim)
  y_sim <- matrix(rnegbin(n = prod(dim(exp_val_sim)), 
                        mu = exp_val_sim, 
                        theta = fit_gam$family$getTheta(TRUE)),
                  nrow = nrow(exp_val_sim), 
                  ncol = ncol(exp_val_sim))
  y_sim<-data.frame(y_sim)%>%
    as_tibble()%>%
    rowid_to_column()%>%
    pivot_longer(!rowid)%>%
    group_by(rowid)%>%
    summarise(value = quantile(value, quants,na.rm=T), q = quants)%>%
    pivot_wider(values_from = value,names_from = q)%>%
    ungroup%>%
    dplyr::select(!rowid)
}

#function to evaluate performance of GAM model (produces daily forecasts)
test_model<-function(series,minyr,leaveout,leave_yrs,Y,S,n){
  for(i in 1:leave_yrs){ 
    last_yr = max(series$year) - (leave_yrs-i+1)
    
    tdat<-series%>%
      mutate(train_test=ifelse(year > last_yr & yday > (S-leaveout), 1, 0))%>%
      filter(year <= (last_yr+1))
    
    m1<-gam(abundance ~  te(yday,year, bs =c("cc","ps"),k=c(10,Y),m=c(2,1)) ,family=nb(theta = NULL, link = "log"), data=tdat%>%filter(train_test==0))
    #gam.check(m1)
    preds<-gam_pred_ints(fit_gam = m1, new_dat = tdat, quants=c(0.025,0.25,0.5,0.75,0.975))%>%
      dplyr::rename(pred = `0.5`,
             `Lo 95` = `0.025`,
             `Hi 95` = `0.975`,
             `Lo 50` = `0.25`,
             `Hi 50` = `0.75`
             )

    tdat<-tdat%>%
      bind_cols(preds)

    p<-ggplot(data=tdat,aes(x=date,y=abundance,group=year,color=factor(train_test)))+
      facet_wrap(~year,scales = "free_x")+
      geom_ribbon(aes(ymin =`Lo 95`, ymax =`Hi 95`), color =NA,fill="cadet blue",alpha=0.5)+
      geom_ribbon(aes(ymin =`Lo 50`, ymax =`Hi 50`), color =NA,fill="cadet blue",alpha=0.5)+
      geom_point()+
      geom_line(mapping = aes(x=date,y=pred),color="cadet blue")+
      theme_bw()

    print(p)
    
    tdat<-tdat%>%
      mutate(predicted_abundance = ifelse(train_test==0, abundance, pred),
             `Lo 95` = ifelse(train_test==0, abundance,`Lo 95`),
             `Hi 95` = ifelse(train_test==0, abundance,`Hi 95`),
             `Lo 50` = ifelse(train_test==0, abundance,`Lo 50`),
             `Hi 50` = ifelse(train_test==0, abundance,`Hi 50`)
             )%>%
      group_by(year)%>%
      summarise(abundance=sum(abundance),
                predicted_abundance=sum(predicted_abundance),
                `Lo 95` = sum(`Lo 95`),
                `Hi 95` = sum(`Hi 95`),
                `Lo 50` = sum(`Lo 50`),
                `Hi 50` = sum(`Hi 50`),
                error = predicted_abundance-abundance
                )%>%
      filter(year==last_yr+1)
    
    if(i==1){results = tdat
    }else{results = results %>% bind_rows(tdat)}
  }
  return(results)
}


#function to evaluate performance of SARIMA model (produces season total forecasts only)
test_model2<-function(series,minyr,leaveout,leave_yrs,Y,S,n){
  for(i in 1:leave_yrs){ 
    last_yr = max(series$year) - (leave_yrs-i+1)
    tdat<-series%>%
      mutate(train_test=ifelse(year > last_yr & yday > (S-leaveout), 1, 0),
             period=ifelse(yday > (S-leaveout), 2,1)
      )%>%
      group_by(year,period,train_test)%>%
      summarize(abundance=sum(abundance),.groups="keep")%>%
      filter(year <= (last_yr+1))
    
    m1<-tdat%>%
      filter(train_test==0)%>%
      ungroup()%>%
      dplyr::select(abundance)%>%
      unlist()%>%
      ts(frequency = 2)%>%
      auto.arima(lambda=0,seasonal = T)
    
    pred<-c(m1$fitted,forecast::forecast(m1,lambda=0,h=1)$mean)
    CI<-forecast::forecast(m1,lambda=0,h=1, level = c(50, 95))%>%
      as_tibble()%>%
      dplyr::select(!`Point Forecast`)%>%
      mutate(year = last_yr+1, period = 2)
      
    
    tdat<-tdat%>%
      bind_cols(pred=pred)%>%
      left_join(CI, by = c("year","period"))
    
    
    p<-ggplot(data=tdat,aes(x=period,y=abundance, group=year,color=factor(train_test)))+
      facet_wrap(~year,scales = "free_x")+
      geom_errorbar(aes(ymin=`Lo 95`, ymax=`Hi 95`), width=.2, color="blue",position=position_dodge(.9))+
      geom_rect(aes(xmin=period-0.25,xmax=period+0.25, ymin=`Lo 50`, ymax=`Hi 50`),fill="white", colour="blue", size=0.5)+ 
      geom_point(mapping = aes(x=period,y=pred),color="blue",shape=3)+
      geom_point(size=2)+
      scale_x_continuous(breaks = c(1,2))+
      theme_bw()+
      ylim(0,NA)

    print(p)
    
    tdat<-tdat%>%
      mutate(predicted_abundance = ifelse(train_test==0, abundance, pred),
             `Lo 95` = ifelse(train_test==0, abundance,`Lo 95`),
             `Hi 95` = ifelse(train_test==0, abundance,`Hi 95`),
             `Lo 50` = ifelse(train_test==0, abundance,`Lo 50`),
             `Hi 50` = ifelse(train_test==0, abundance,`Hi 50`)
             )%>%
      group_by(year)%>%
      summarise(abundance=sum(abundance),
                predicted_abundance=sum(predicted_abundance),
                `Lo 95` = sum(`Lo 95`),
                `Hi 95` = sum(`Hi 95`),
                `Lo 50` = sum(`Lo 50`),
                `Hi 50` = sum(`Hi 50`),
                error = predicted_abundance-abundance
                )%>%
      filter(year==last_yr+1)
    
    if(i==1){results = tdat
    }else{results = results %>% bind_rows(tdat)}
  }
  return(results)
}

```

Here we will load & install packages we need to use (needs internet connection if packages not already installed)
```{r load_packages, message = FALSE, warning = FALSE,results = "hide"}
packages_list<-c("tidyverse"
                 ,"forecast"
                 ,"mgcv"
                 ,"ggplot2"
                 ,"MASS"
                 ,"RColorBrewer"
                 ,"kableExtra"
                 ,"lubridate"
                 ,"modelr"
                 ,"kableExtra"
                 ,"reshape2"
                 ,"ggfortify"
                 ,"clock"
                 ,"smooth"
                 )
install_or_load_pack(pack = packages_list)
```

## User Inputs

```{r user_inputs, message = FALSE, warning = FALSE,results = "show"}
#===========================
#simulated timeseries params
#===========================
S<-55 #days in a season (must currently be < 59)
minyr<-2010 #first year in simulated dataset
Y<-20 #years to simulate
Y_resid<-cumsum(rnorm(Y,0,0.5)) #changes to annual abundance
n<-Y*S #sample size of seasonal timeseries to simulate
phi<-c(0.2,0.4) #AR terms (lag1, seasonal lag 1)
theta<-c(0.2,0.5) #MA terms (lag1, seasonal lag 1)
proc_sd<-0.25 #process error sd for run timing evolution
int<-4 #log mean daily abundance
#==================
#forecasting params
#==================
leaveout<-35 # days of season to leave out of forecast
leave_yrs<- 10
```

## Simulate Data
```{r Analysis_simulate_data, message=FALSE, warning=FALSE, results="show"}
series<-make_data(S,min_yr,Y,Y_resid,n,phi,theta,proc_sd)
print(series)

p<-ggplot(data=series,aes(x=date,y=abundance,group=year))+
  facet_wrap(~year,scales = "free_x")+
  geom_point()+
  geom_line()
print(p)

ggplot(series%>%group_by(year)%>%summarise(abundance=sum(abundance),.groups = "keep"), aes(x=year,y=abundance))+
  geom_line()+
  geom_point()
```

## GAM results

```{r Analysis_v1, message=FALSE, warning=FALSE, results="show"}
results<-test_model(series=series,minyr=minyr,leaveout=leaveout,leave_yrs=leave_yrs,Y=Y,S=S,n=n)
results<-results%>%
  mutate(pct_error=(error/abundance)*100)

results%>%
  kbl(caption = "Table 1.GAM Results",digits =0)%>%
  kable_classic(full_width = F, html_font = "Cambria")

print(paste0("MAPE = ",mean(abs(results$pct_error))))

```

## SARIMA results

```{r Analysis_v2, message=FALSE, warning=FALSE, results="show"}
results2<-test_model2(series=series,minyr=minyr,leaveout=leaveout,leave_yrs=leave_yrs,Y=Y,S=S,n=n)
results2<-results2%>%
  mutate(pct_error=(error/abundance)*100)

results2%>%
  kbl(caption = "Table 2.SARIMA Results",digits =0)%>%
  kable_classic(full_width = F, html_font = "Cambria")

print(paste0("MAPE = ",mean(abs(results2$pct_error))))

      
```
